---
title: "ACS Table Pipeline"
author: "United Way Team"
date: "2024-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Needed packages
library(tidyverse)
library(knitr)
library(tidycensus)

# Census API key
census_api <- Sys.getenv("CENSUS_API_KEY")
```

## Step 1: Determining Variables of Interest

We begin by determining the variables we are interested in pulling data, the following list is the variables and their respective sub-categories, we want to pull from the ACS.

- DP05: Race and Ethnicity %
  - DP05_0038P: Black/AA
  - DP05_0039P: American Indian/Alaska Native
  - DP05_0044P: Asian
  - DP05_0052P: Native Hawaiian/Pacific Islander
  - DP05_0071P: Hispanic/Latino (any race)
  - DP05_0077P: White alone (not Hispanic/Latino)
  - DP05_0033P: All
- S0901: Race and Ethnicity of Children %
  - S0901_C01_007: Black/AA
  - S0901_C01_008: American Indian/Alaska Native
  - S0901_C01_009: Asian
  - S0901_C01_010: Native Hawaiian/Pacific Islander
  - S0901_C01_013: Hispanic/Latino (any race)
  - S0901_C01_014: White alone (not Hispanic/Latino)
  - S0901_C01_001: All
- S1701: Poverty Rates
  - S1701_C02_014: Black/AA
  - S1701_C02_015: American Indian/Alaska Native
  - S1701_C02_016: Asian
  - S1701_C02_017: Native Hawaiian/Pacific Islander
  - S1701_C02_020: Hispanic/Latino (any race)
  - S1701_C02_021: White alone (not Hispanic/Latino)
  - S1701_C02_001: All
- S1903: Median Household Income
  - S1903_C03_003: Black/AA
  - S1903_C03_004: American Indian/Alaska Native
  - S1903_C03_005: Asian
  - S1903_C03_006: Native Hawaiian/Pacific Islander
  - S1903_C03_009: Hispanic/Latino (any race)
  - S1903_C03_010: White alone (not Hispanic/Latino)
  - S1903_C03_001: All
- S2701: Without Health Insurance %
  - S2701_C03_017: Black/AA
  - S2701_C03_018: American Indian/Alaska Native
  - S2701_C03_019: Asian
  - S2701_C03_020: Native Hawaiian/Pacific Islander
  - S2701_C03_023: Hispanic/Latino (any race)
  - S2701_C03_024: White alone (not Hispanic/Latino)
  - S2701_C03_001: All
- C27001 & B18135: Children without Health Insurance %
  - C27001B_004: Black/AA
  - C27001C_004: American Indian/Alaska Native
  - C27001D_004: Asian
  - C27001E_004: Native Hawaiian/Pacific Islander
  - C27001I_004: Hispanic/Latino (any race)
  - C27001H_004: White alone (not Hispanic/Latino)
  - : All
- S2301: Unemployment Rate %
  - S2301_C04_013: Black/AA
  - S2301_C04_014: American Indian/Alaska Native
  - S2301_C04_015: Asian
  - S2301_C04_016: Native Hawaiian/Pacific Islander
  - S2301_C04_019: Hispanic/Latino (any race)
  - S2301_C04_020: White alone (not Hispanic/Latino)
  - S2301_C04_001: All
- S2502: Homeownership % 
  - S2502_C02_003: Black/AA
  - S2502_C02_004: American Indian/Alaska Native
  - S2502_C02_005: Asian
  - S2502_C02_006: Native Hawaiian/Pacific Islander
  - S2502_C02_009: Hispanic/Latino (any race)
  - S2502_C02_010: White alone (not Hispanic/Latino)
  - S2502_C02_001: All

We can create vector lists of these variables, we can then pull the variables from the table containing all the metadata:

```{r}
# Create vector of variables of interest
# # Race and Ethnicity %
var_of_interest <- list(
  "DP05_0038P" = "RaceEtnic_%_Black", # Black/AA
  "DP05_0039P" = "RaceEtnic_%_AmerIndian", # American Indian/Alaska Native
  "DP05_0044P" = "RaceEtnic_%_Asian", # Asian
  "DP05_0052P" = "RaceEtnic_%_PacifIslan", # Native Hawaiian/Pacific Islander
  "DP05_0071P" = "RaceEtnic_%_HispanLatin", # Hispanic/Latino (any race)
  "DP05_0077P" = "RaceEtnic_%_White", # White alone (not Hispanic/Latino)
  "DP05_0033P" = "RaceEtnic_%_All", # All
  # Race and Ethnicity of Children %
  "S0901_C01_007" = "RaceEthnicChild_%_Black", # Black/AA
  "S0901_C01_008" = "RaceEthnicChild_%_AmerIndian", # American Indian/Alaska Native
  "S0901_C01_009" = "RaceEthnicChild_%_Asian", # Asian
  "S0901_C01_010" = "RaceEthnicChild_%_PacifIslan", # Native Hawaiian/Pacific Islander
  "S0901_C01_013" = "RaceEthnicChild_%_HispanLatin", # Hispanic/Latino (any race)
  "S0901_C01_014" = "RaceEthnicChild_%_White", # White alone (not Hispanic/Latino)
  "S0901_C01_001" = "RaceEthnicChild_%_All", # All
  # Poverty Rates
  "S1701_C02_014" = "PovertyRate_Black", # Black/AA
  "S1701_C02_015" = "PovertyRate_AmerIndian", # American Indian/Alaska Native
  "S1701_C02_016" = "PovertyRate_Asian", # Asian
  "S1701_C02_017" = "PovertyRate_PacifIslan", # Native Hawaiian/Pacific Islander
  "S1701_C02_020" = "PovertyRate_HispanLatin", # Hispanic/Latino (any race)
  "S1701_C02_021" = "PovertyRate_White", # White alone (not Hispanic/Latino)
  "S1701_C02_001" = "PovertyRate_All", # All
  # Median Household Income
  "S1903_C03_003" = "MedianHouseIncome_Black", # Black/AA
  "S1903_C03_004" = "MedianHouseIncome_AmerIndian", # American Indian/Alaska Native
  "S1903_C03_005" = "MedianHouseIncome_Asian", # Asian
  "S1903_C03_006" = "MedianHouseIncome_PacifIslan", # Native Hawaiian/Pacific Islander
  "S1903_C03_009" = "MedianHouseIncome_HispanLatin", # Hispanic/Latino (any race)
  "S1903_C03_010" = "MedianHouseIncome_White", # White alone (not Hispanic/Latino)
  "S1903_C03_001" = "MedianHouseIncome_All", # All
  # Without Health Insurance %
  "S2701_C03_017" = "NoHealthInsur_%_Black", # Black/AA
  "S2701_C03_018" = "NoHealthInsur_%_AmerIndian", # American Indian/Alaska Native
  "S2701_C03_019" = "NoHealthInsur_%_Asian", # Asian
  "S2701_C03_020" = "NoHealthInsur_%_PacifIslan", # Native Hawaiian/Pacific Islander
  "S2701_C03_023" = "NoHealthInsur_%_HispanLatin", # Hispanic/Latino (any race)
  "S2701_C03_024" = "NoHealthInsur_%_White", # White alone (not Hispanic/Latino)
  "S2701_C03_001" = "NoHealthInsur_%_All", # All
  # Children without Health Insurance %
  "C27001B_004" = "NoHealthInsurChild_%_Black", # Black/AA
  "C27001C_004" = "NoHealthInsurChild_%_AmerIndian", # American Indian/Alaska Native
  "C27001D_004" = "NoHealthInsurChild_%_Asian", # Asian
  "C27001E_004" = "NoHealthInsurChild_%_PacifIslan", # Native Hawaiian/Pacific Islander
  "C27001I_004" = "NoHealthInsurChild_%_HispanLatin", # Hispanic/Latino (any race)
  "C27001H_004" = "NoHealthInsurChild_%_White", # White alone (not Hispanic/Latino)
  "NA" = "NoHealthInsurChild_%_All", # All
  # Unemployment Rate %
  "S2301_C04_013" = "UnemployRate_%_Black", # Black/AA
  "S2301_C04_014" = "UnemployRate_%_AmerIndian", # American Indian/Alaska Native
  "S2301_C04_015" = "UnemployRate_%_Asian", # Asian
  "S2301_C04_016" = "UnemployRate_%_PacifIslan", # Native Hawaiian/Pacific Islander
  "S2301_C04_019" = "UnemployRate_%_HispanLatin", # Hispanic/Latino (any race)
  "S2301_C04_020" = "UnemployRate_%_White", # White alone (not Hispanic/Latino)
  "S2301_C04_001" = "UnemployRate_%_All", # All
  # Homeownership % 
  "S2502_C02_003" = "Homeown_%_Black", # Black/AA
  "S2502_C02_004" = "Homeown_%_AmerIndian", # American Indian/Alaska Native
  "S2502_C02_005" = "Homeown_%_Asian", # Asian
  "S2502_C02_006" = "Homeown_%_PacifIslan", # Native Hawaiian/Pacific Islander
  "S2502_C02_009" = "Homeown_%_HispanLatin", # Hispanic/Latino (any race)
  "S2502_C02_010" = "Homeown_%_White", # White alone (not Hispanic/Latino)
  "S2502_C02_001" = "Homeown_%_All" # All
  ) 
```

```{r}
# Create function to get all metadata in single file
all_acs_meta <- function(){
  # Gets the list of all variables from all acs5 metadata tables
  vars1 <- load_variables(2021, "acs5") %>% select(-geography) # Remove the geography column
  vars2 <- load_variables(2021, "acs5/profile")
  vars3 <- load_variables(2021, "acs5/subject")
  vars4 <- load_variables(2021, "acs5/cprofile")
  
  # Provides column with specific lookup
  vars1$dataset_table <- "acs5"
  vars2$dataset_table  <- "acs5/profile"
  vars3$dataset_table  <- "acs5/subject"
  vars4$dataset_table  <- "acs5/cprofile"
  
  # Combine all table rows
  all_vars_meta <- rbind(vars1, vars2, vars3, vars4)
  
  return(all_vars_meta)
}

# Pull all variable names from metadata
metadata_var <- all_acs_meta()
```


Next, we create a table of the metadata for the variables, we will join to this list after we pull in the data using the API:

```{r}
# Filters for only the rows containing the variable codes we are interested in
var_table <- metadata_var %>% 
  filter(name %in% names(var_of_interest))
```

We can print the output of the table below:

```{r, echo=FALSE}
# Prints out the table in 
kable(var_table)
```



## Step 2: Acquire Data from Census Bureau Tables

Next we want to take each variable and pull it for all of our locations and years:

Once again, we should determine our regions and years.
- Our years of interest can be input as a simple vector list
- Our regions of interest can be input as a 3-column table.
  1. Name of the region
  2. Code for the region (three-digit FIPS county code)
  3. Combined region grouping (counties that share a region have the same group value)
    - Greater Charlottesville Area (GCA): Albemarle, Charlottesville, Fluvanna, Greene, Louisa, Nelson
    - Staunton/Augusta/Waynesboro (SAW): Staunton, Augusta, Waynesboro

```{r}
# Create list of years
yr_of_interest <- c(2019, 2020)

# Create list of county names
Name = c("Albemarle", "Charlottesville", "Fluvanna", "Greene", "Louisa", "Nelson", "Staunton", "Augusta", "Waynesboro",
       "Harrisonburg", "Buckingham", "Orange", "Rockingham", "Madison", "Page", "Rappahannock", "Shenandoah", "Warren", 
       "Winchester", "Clarke", "Frederick")

# Create list of county FIPs code
Code = c("003", "540", "065", "079", "109", "125", "790", "015", "820", "660", "029", "137", "165", "113", "139", "157", "171", "187", "840", "043", "069")

# Create list of county groupings
Group = c(rep("GCA", 6), rep("SAW", 3), "Harrisonburg", "Buckingham", "Orange", "Rockingham", "Madison", "Page", "Rappahannock",
          "Shenandoah", "Warren", "Winchester", "Clarke", "Frederick")
  
# Create table of regions
region_of_interest <- tibble(Name, Code, Group)
```

Then, we can use the `get_acs()` function from the `tidycensus` package to pull all these variables for each of these localitites and years:

```{r}
# Pulls from census API using tidycensus
pulled_data <- get_acs(geography = "county",
        variables = names(var_of_interest), # Variables list provided in the all_acs5_data function wrapping this one
        state = "51", # Virginia code
        county = Code, 
        year = 2019, 
        survey = "acs5",
        output = "wide",
        key = census_api) # API key provided in the all_acs5_data function
```

From here we can link the rows representing county regions with our datframe from earlier:

```{r}
# Adds 51 to beginning of every 3-digit FIPS code to match with the pulled data
region_of_interest <- region_of_interest %>% mutate(Code=paste0("51", Code))

# Join the regions dataframe to the pulled data
pulled_data_full <- merge(x=region_of_interest[,2:3], y=pulled_data, by.x = "Code", by.y = "GEOID")

# Rename columns 
colnames(pulled_data_full)[1:3] <- c("GEOID", "Group", "Region")
```

Now we take the column names and give them readable labels instead of codes:

```{r}
# Create list of the old column names
old_col_names <- c(paste0(names(var_of_interest), "E"), paste0(names(var_of_interest), "M"))
# Gives list of the new column names
new_col_names <- c(paste0(unname(unlist(var_of_interest)), "E"), paste0(unname(unlist(var_of_interest)), "M"))

pulled_data_full <- data.table::setnames(pulled_data_full, old = old_col_names, new = new_col_names, skip_absent = T)
```




## Step 3: Aggregating Table Data for Combined Regions

We can now group by the region column and produce aggregate values:

```{r}
# Group margin of error columns by shared region
pulled_data_full_agg2 <- pulled_data_full %>%
  group_by(Group) %>%
  summarise(across(ends_with("E"), ~sum(.)),
            across(ends_with("M"), ~moe_sum(., estimate=sub(".$", "E", .))))
```


